Variance reduction for MCMC methods via martingale representations

### 2020-02-20 aap/Revision/vr-mcmc-revision-new.tex
update

\section{Numerical results}
\label{sec:num}
If each function  \(Q_{p,l}(x) \) can be well approximated by polynomials, then we can use polynomial basis functions in   \eqref{eq:06042018a1} to approximate \(\bar Q_l\). Suppose that we constructed a polynomial approximation  for each function \(\bar Q_{l}\)  in the form:
\widehat{Q}_{l}(x) = \sum_{\|\mathbf{s}\|\leq m} \beta_{\mathbf{s}} x^{\mathbf{s}},\quad s=(s_1,\ldots,s_d)
with some coefficients \(\beta_{\mathbf{s}}\in \mathbb{R}.\)
Then using the identity
\xi^j = j! \sum_{r = 0}^{\lfloor j/2 \rfloor} \frac{1}{2^r  r! \sqrt{(j-2r)!}} H_{j-2r}(\xi),\quad \xi \in \mathbb{R},
we derive for all $x \in \mathbb{R}^d$,
\widehat a_{l,\mathbf{k}} (x) &=&\CPE{\mathbf{H}_\mathbf{k}(x) \widehat{Q}_{l}(x - \gamma_l \mu(x) + \sqrt{\gamma_l}\xi)}{\TrainSet \vee \mathcal{G}_N}
= \sum_{\|\mathbf{s}\|\leq m} \beta_{\mathbf{s}} \prod_{i=1}^d P_{i,k_i,s_i}(x),
where for all integers $i,k,s_i$ and $x \in \rset^d$,
P_{l,k,s}(x) &= \mathsf{E} \left[ H_{k}(\xi_l) (x_l - \gamma_l \mu_l(x) + \sqrt{\gamma_l} \xi_l)^{s} \right]
is a one-dimensional polynomial (in \(x\)) of degree at most \(s\) with analytically known coefficients. Alternatively, on practice we can work directly with functions \(Q_{p,l}.\) Due to (almost) time-homogenous Markov chain generated by ULA, each function $Q_{p,l}$ can be approximated by function, which depends only on the  $p-l$, that is,
 Q_{p,l}(x) = \mathsf{E} \left[f(X_p) | X_l = x\right] \approx Q^\circ_{p-l}(x)
Thus we have
  Q^\circ_{r}(x) = \mathsf{E} \left[f(X_{l+r}) | X_l = x \right].
Consequently, the functions $Q^\circ_r$ can be estimated using a modified least squares criteria
\label{eq:Qcircr}
\widehat{Q}^\circ_r = \argmin_{\psi \in \Psi} \sum_{s = 1}^{T} \sum_{l = N + 1}^{N+n-r} \left| f(X_{l+r}^{(s)}) - \psi(X_l^{(s)})\right|^2
for \(1 \leq r \leq n-1,\) where  $\widehat{Q}^\circ_0(x) = f(x)$ by definition.
Due to Lemma~\ref{a_decay}, it is enough to estimate $Q^\circ_{r}$ for $|r| < n_{\mathrm{trunc}}$ for some truncation level \(n_{\mathrm{trunc}}\) depending on \(d\) and \(\gamma.\) It allows us to use a smaller amount of training trajectories to approximate conditional expectations $Q^\circ_{r}$.
Finally  we construct a truncated version of our estimator:
\pi_{K,n,n_{\mathrm{trunc}}}^N(f) = \pi_n^N(f) -  M_{K,n, n_{\mathrm{trunc}}}^{N}(f),
{\widehat  M}_{K,n, n_{\mathrm{trunc}}}^N &=& \sum_{0<\left\| \mathbf{k}\right\| \leq K} \sum_{p=N+1}^{N+n} \omega_{p,n}^N
&&\times
\sum_{l=N+1}^p {\widehat  a}_{p-l,\mathbf{k}}(X_{l-1})\mathbf{H}_{\mathbf{k}}(Z_l) \mathbbm{1}\{ |p-l| < n_{\mathrm{trunc}} \}
and
\widehat a_{p-l,\mathbf{k}} (x) &=&\CPE{\mathbf{H}_\mathbf{k}(x) \widehat{Q}^\circ_{p-l}(x - \gamma_l \mu(x) + \sqrt{\gamma_l}\xi)}{\TrainSet \vee \mathcal{G}_N}.
\subsection{Gaussian mixtures}
 We consider a sample generated by ULA with $\pi$  given by the mixture of two Gaussian distributions  with equal weights:
\pi (x) = \frac{1}{2(2\pi)^{d/2}} \left( \rme^{\frac{-\| x-a\|_2^2}{2}}  + \rme^{\frac{-\| x+a\|_2^2}{2}} \right), \quad x \in\mathbb R^d
where $a \in \mathbb{R}^d$ is a given vector. The function $U(x)$ and its gradient are given by
U(x) = \frac{1}{2} \|x - a\|_2^2 - \text{log}(1 + \rme^{-2x^\top a}), \,\,
\nabla U(x) = x-a +2a(1 + \rme^{2 x^\top a})^{-1},
respectively.
In our experiments we considered dimensions $d = 2$ and $d = 8$ and take $a=( (2d)^{-1/2}, \dots, (2d)^{-1/2})^\top$.
In order to approximate the expectation \(\pi(f)\) with \(f(x)=\sum_{i=1}^d x_i,\)
 we have used constant step sizes $\gamma_i = \gamma=0.2$ and sampled $n = 10^4$ samples for $d = 2$ and $n = 2 \times 10^3$ for $d = 8$.  We generated $\NtrainPath = 10$ independent "training" trajectories and solved the least squares  problems \eqref{eq:Qcircr} using the first order polynomial approximations for the coefficients \(a_{p,\mathbf{k}}\) as described in the previous section. The truncation level \(n_{\mathrm{trunc}}\) is chosen to be \(50.\) To test our variance reduction algorithm, we generated 100 independent trajectories.
 In Figure ~\ref{fig:1} we compare our approach to variance reduction methods of \cite{mira2013zero} and \cite{brosse2018diffusion}.

>>>>

\section{Numerical results}
\label{sec:num}
We will perform numerical experiments based on ULA with constant step size $\gamma$. Suppose that each function  \(Q_{r}(x)\) from \ref{eq:definition-bar-a} can be well approximated by polynomials, and we constructed a polynomial approximation  for each \(Q_r(x)\)  in the form:
\begin{equation*}
\widehat{Q}_{r}(x) = \sum_{\|\mathbf{s}\|\leq m} \beta_{\mathbf{s}} x^{\mathbf{s}},\quad s=(s_1,\ldots,s_d)
\end{equation*}
with some coefficients \(\beta_{\mathbf{s}}\in \mathbb{R}.\)
Then using the identity
\[
\xi^j = j! \sum_{r = 0}^{\lfloor j/2 \rfloor} \frac{1}{2^r  r! \sqrt{(j-2r)!}} H_{j-2r}(\xi),\quad \xi \in \mathbb{R},
\]
we obtain closed-form expression for the estimates $\widehat a_{r,k}(x)$ of functions $\bar a_{r,k}(x)$ from \ref{eq:definition-bar-a}. Namely, for all $x \in \mathbb{R}^d$, 
\begin{eqnarray*}
\widehat a_{r,\mathbf{k}} (x) &=&\CPE{\mathbf{H}_\mathbf{k}(x) \widehat{Q}_{r}(x - \gamma \mu(x) + \sqrt{\gamma}\xi)}{\TrainSet \vee \mathcal{G}_N}
= \sum_{\|\mathbf{s}\|\leq m} \beta_{\mathbf{s}} \prod_{i=1}^d P_{i,k_i,s_i}(x),
\end{eqnarray*}
where for all integers $i,k_i,s_i$ and $x \in \rset^d$,
\begin{align*}
P_{r,k,s}(x) &= \mathsf{E} \left[ H_{k}(\xi_l) (x - \gamma \mu(x) + \sqrt{\gamma} \xi)^{s} \right]
\end{align*}
is a one-dimensional polynomial (in \(x\)) of degree at most \(s\) with analytically known coefficients. In consequence, it is enough to estimate functions \(Q_{r}(x)\). It can be done using a modified least-squares criteria based on $T$ training trajectories \(\left(X_1^{(s)},\ldots,X_{N+n}^{(s)}\right), s = 1,\ldots,T\) with $N$ being the size of burn-in period:
\begin{eqnarray}
\label{eq:Qcircr}
\widehat{Q}_r = \argmin_{\psi \in \Psi} \sum_{s = 1}^{T} \sum_{l = N + 1}^{N+n-r} \left| f(X_{l+r}^{(s)}) - \psi(X_l^{(s)})\right|^2
\end{eqnarray}
for \(1 \leq r \leq n-1,\), $\Psi$ being the class of polynomials $\Psi = \{\psi(x) | \psi(x) = \sum_{\|\mathbf{s}\|\leq m} \alpha_{\mathbf{s}} x^{\mathbf{s}} \}$ and $\widehat{Q}_0(x) = f(x)$ by definition. Due to Lemma~\ref{a_decay},
\sergey{we need to put some lemma about the truncation point here, I will do it}
it is enough to estimate $Q_{r}$ for $r < n_{\mathrm{trunc}}$ for some truncation level \(n_{\mathrm{trunc}}\) depending on \(d\) and \(\gamma.\) It allows us to use a smaller amount of training trajectories to approximate $Q_{r}$.

Finally  we construct a truncated version of the estimator from \ref{}:
\[
\widehat\pi_{K,n,n_{\mathrm{trunc}}}^N(f) = \widehat\pi_n^N(f) -  \widehat M_{K,n, n_{\mathrm{trunc}}}^{N}(f),
\]
where
\begin{eqnarray*}
\widehat M_{K,n,n_{\mathrm{trunc}}}^N(f) &=&\frac{1}{n}\sum_{p=N+1}^{N+n}\left[\sum_{k=1}^{K}\sum_{l=N+1}^{p} \bar a_{p-l,k}(X_{l-1})\phi_k(\xi_{l})\right]
\end{eqnarray*}
and
\begin{eqnarray*}
\widehat a_{p-l,\mathbf{k}} (x) &=&\CPE{\mathbf{H}_\mathbf{k}(x) \widehat{Q}^\circ_{p-l}(x - \gamma_l \mu(x) + \sqrt{\gamma_l}\xi)}{\TrainSet \vee \mathcal{G}_N}.
\end{eqnarray*}

\subsection{Gaussian mixtures}
 We consider a sample generated by ULA with $\pi$  given by the mixture of two Gaussian distributions  with equal weights:
\[
\pi (x) = \frac{1}{2(2\pi)^{d/2}} \left( \rme^{\frac{-\| x-a\|_2^2}{2}}  + \rme^{\frac{-\| x+a\|_2^2}{2}} \right), \quad x \in\mathbb R^d
\]
where $a \in \mathbb{R}^d$ is a given vector. The function $U(x)$ and its gradient are given by
\[
U(x) = \frac{1}{2} \|x - a\|_2^2 - \text{log}(1 + \rme^{-2x^\top a}), \,\,
\nabla U(x) = x-a +2a(1 + \rme^{2 x^\top a})^{-1},
\]
respectively.
In our experiments we considered dimensions $d = 2$ and $d = 8$ and take $a=( (2d)^{-1/2}, \dots, (2d)^{-1/2})^\top$.
In order to approximate the expectation \(\pi(f)\) with \(f(x)=\sum_{i=1}^d x_i,\)
 we have used constant step sizes $\gamma_i = \gamma=0.2$ and sampled $n = 10^4$ samples for $d = 2$ and $n = 2 \times 10^3$ for $d = 8$.  We generated $\NtrainPath = 10$ independent "training" trajectories and solved the least squares  problems \eqref{eq:Qcircr} using the first order polynomial approximations for the coefficients \(a_{p,\mathbf{k}}\) as described in the previous section. The truncation level \(n_{\mathrm{trunc}}\) is chosen to be \(50.\) To test our variance reduction algorithm, we generated 100 independent trajectories.
 In Figure ~\ref{fig:1} we compare our approach to variance reduction methods of \cite{mira2013zero} and \cite{brosse2018diffusion}.
